Glyph-maps for visualizing climate data and models

### 2011-09-27 paper.tex
Changes to first two sections

Climate data is composed of measurements on variables such as temperature, precipitation, and winds that have been collected from a spatial and temporal context. The classic display for spatial data is to use color or grey scale overlaid on geographic location, typically including a map. When measurements are made at different time periods it is common to display the data as small multiples, such as in Figure \ref{fig:facetted-map}. In this plot, a separate map is drawn for each month (columns) and year (rows), over a 6 year period of remotely sensed temperature data above Central America. Color is used to display the de-seasonalized temperature measurements, with red being high values and blue being low values. The data comes from \citet{murrell:2010}. The most noticeable feature in this display is the strong red in the equatorial region beginning in mid-1997 and tapering out during 1998. This is the pattern of an El Ni\~no event, a major temperature anomaly. With some work other smaller more localized patterns can be perceived, for example, cooler over the land mass in the early year, but warmer in the latter years. What the reader is required to do is play the game ``Spot the difference'' from one small image to another. From a cognitive perspective, this is a very hard task to do. It falls into the framework of {\it change blindness} (see e.g. \citet{healey:2011} and often leads to failure without readers realizing, in particular for small areas of change or subtle changes.  Large structures such as the El Ni\~no event are clear but readers will not be able to effectively glue the images together sufficiently to read off the trend over the years or notice small local deviations between plots. Rendering these small multiples as movie helps with picking up strong trends, but small trends still fail to draw readers' attention and escape unnoticed \citep{simons:gradual}.

\begin{figure*}[htp]
\centerline{\includegraphics[width=6in]{nasa-colored-map.png}}
\caption{Facetted map of de-seasonalized temperature. The dominant feature is the El Ni\~no warming in the southern equatorial region in the last half of 1997 and first half of 1998. Smaller features are only noticeable on closer inspection, or if pointed out: such as the relative warming on the mountain regions in south and north America in the later years.}
\label{fig:facetted-map}
\end{figure*}

@@

In this paper we explore the use of a different type of display, that uses icons on the map, for climate data.  Icon plot associated with weather maps date back as far as \citet{galton:weathermap}, described in \citet{friendlydenis:2001}. Icon plots date back to the star plot introduced by \citet{mayr:1877}, and resurged in the modern era with the semi-graphic displays of \citet{anderson:1960}, and were made infamous by \citep{chernoff:1973}. An icon is produced by mapping each variable to some feature of the icon. One icon is generated for every case in the data set. %Figure \ref{fig:glyphs} illustrates the construction of star plot, with emphasis on temporal data. 
For temporal data, time measurements can be treated as variables in a multivariate data set, fitting this into the icon plot framework. Figure \ref{fig:templates} displays 12 iconic time series shapes, and how they would look if the icon was a time series, or displayed as a star glyph. The data underlying each glyph has 36 time variables. The star glyphs is formed by considering the 36 axes radiating from a common midpoint, and the data values for the row are plotted on each axis relative to the locations of the minimum and maximum of the variable. For the temporal data, all variables are considered to be on a common scale so the global minimum and maximum of temperature are used to scale each data value. 
%Three different locations are selected, for demonstration purposes. The top plot shows the time series of these three locations. Location ``1.1'' (red) has medium values of temperature, with strong seasonality, the six years corresponding to the six peaks. Location ``2.1'' (green) has generally high temperature values, with weaker seasonality in most years and a lack of low temperatures in the middle of the time period when lows would be expected. Location ``3.1'' (blue) has strong seasonality but generally low temperatures. The bottom plot shows these same cases as star glyphs. A separate glyph is created for each location. The 72 time variables form axes organized radially. Each temperature value is plotted relative to the overall range of temperature values. The seasonal patterns are clearly visible in this type of display, forming six ``petals'' in the icon. Location ``2.1'' has a different pattern than the other two, with only four petals, and location ``3.1'' is a much smaller icon indicating overall lower values.

@@

%\begin{figure*}[htp]
%\centerline{\includegraphics[width=6in]{nasa-glyph-ts.pdf}}
%\centerline{\includegraphics[width=6in]{nasa-glyph.pdf}}
%\caption{Icons of raw temperature for three different locations.}
%\label{fig:glyphs}
%\end{figure*}

Icon plots are used, infrequently, to examine multivariate data. For large data sets it is infeasible to digest the quantity of icons, and for multivariate data there is no natural order so it can be like having a mountain of little stamps to rearrange and organize, in order to understand the structures present in the data. An early paper on multivariate graphics \citep{kleiner:1981} and a recent paper \citep{hurley:2010} propose some approaches for organizing the icons. Ordering of icons is not an issue with spatiotemporal data, because it is naturally handled by spatial location. The spatial trend, or continuity, usually gives the plot an appearance of texturing on a landscape, make it easier to digest patterns among the icons.

In this paper we describe the algorithm for creating glyphs of temporal data on maps, generally in R, and the various modifications that can enable different features in the data to be detected. The next section (\ref{sec:glyph-map}) describes the use of glyphs on maps, and Section \ref{sec:construction} explains how to construct glyph-maps. Section \ref{sec:reference} discusses the use of reference lines and boxes, and Section \ref{sec:scale} visits issues of scaling values for different information purposes. Generalizations to different types of icons are described in Section \ref{sec:scatter}. Glyph-maps work nicely with gridded data, but many spatiotemporal data sets have irregular spatial locations. Possible adjustments are discussed in Section \ref{sec:irregular}. Code and data is available in the online supplementary materials.

\section{Glyph-maps}~\label{sec:glyph-map}

% Reference Tufte's small multiples
\emph{Glyph-maps} are maps where a glyph is displayed at each spatial location to represent multiple data values. To explain the glyph-map, we use data of 2$^o$x2$^o$ gridded temperature anomalies, from the GISTEMP surface temperature data provided by the Earth System Research Laboratory, Physical Sciences Division, National Oceanic and Atmospheric Administration. Ground station data was de-seasonalized, differenced from ***, and spatially averaged to obtain gridded measurements. Figure \ref{fig:gistemp-pred} shows the time series icons for each spatial location of predicted temperature for 1950-2010 across the USA. Most locations observe an increasing temperature, to varying degrees, over this time period. A few locations show a decline. These locations tend to be different from their neighbors, which might suggest isolated data problems. There is a difference in the temperature increase at different regions: steeper inclines can be seen in the north and western mountain region and across the Great Lakes, but inclines are flatter in the southeastern USA. 
%The most seasonality occurs in north American. Some seasonal effects can also be seen in the Caribbean and in south American locations. Around the equator the temperatures are relatively warm for all seasons. Instead of radially displaying the temperature a glyph can be a regular time series, time displayed horizontally and temperature vertically (right plot). The structure perceived is different with this change in glyph type: locations at the equator have flat series, on land more varied temperatures.

\begin{figure}[htbp]
  \centering
  \includegraphics[width=1\linewidth]{gistemp-pred}%

  %\includegraphics[width=1\linewidth]{gistemp-polar-pred}

% Add time series as well as radial
%\begin{figure*}[htp]
%\centerline{\includegraphics[width=3.5in]{nasa-glyph-all.pdf}\includegraphics[width=3.5in]{nasa-glyph-ts-all.pdf}}
%\caption{Glyph-map of raw data, using star glyphs (left), and time series as glyphs (right).}
%\label{fig:glyph-map}
%\end{figure*}

Figure \ref{fig:gistemp-pred} shows the star glyphs for each spatial location of raw temperature for 1880-2011 across the USA. The major feature to notice is the pattern of missing values, icons that do not complete the full circle. The high noon position of the circle is the first time point, Jan 1880, and it is this early time period that has missing values are a lot of locations. It was for this reason that the temporal model was fit just to 1950-2010 data. Another important feature is variability, indicated by the thickness of the circle. Some locations, particularly in the north western mountain region, have big variability in temperature, both low minima, and high maxima. At these locations there are large fluctuations from the long term mean.

@@

Icons have been displayed on maps since the earliest displays weather forecasts, in the 1800s. In recent times, \citet{pickett:1988} introduced the idea of using icons on maps to display multivariate spatial data. This area of work has morphed into the field of metaphorical data displays, which creates abstract landscapes of spatial data, a digression from glyph-maps. \citet{gribov:2006} describes the use of glyph-maps for multivariate data also, with emphasis on the graphics software Gauguin.

%\subsection{Glyph plots}

%Avoids one problem that plagues glyph displays: the ordering of the variables. Although some techniques exist to ameliorate this problem \citep{kleiner:1981,hurley:2010}, they are not necessary with time data because it is intrinsically ordered.

%Will first explore for regularly gridded data, a simple starting point and a common output of climate models, and then show a few simple extensions to deal with non-rectangular grids and irregular locations, as exemplified by raw climate data which is typically collect at irregular locations.

@@

%A big advantage of glyph-maps is that they can be generated with existing graphics software after performing a simple pre-processing step, making them easily accessible to a wide audience. 
Creating a glyph-map, involves recognizing that they combine two structural components of the data: spatial location and data values. The spatial location is the major positioning component, and the data values are minor adjustments to these positions. For spatiotemporal data, the major axes are latitude ($y_{\amaj}$) and longitude ($x_{\amaj}$), and the minor axes are time ($x_{\amin}$) and some measurement ($y_{\amin}$), for example, temperature, or predicted temperature. The minor axes are rescaled to have range $[0, 1]$, and further adjusted to fit into the resolution of the spatial coordinates. The final values to be plotted are a simple linear combination, as given by

\begin{equation}
\begin{array}{lll}
x_\text{ coordinate}&=& x_{\amaj} + w \cdot x_{\amin}\\
y_\text{ coordinate}&=& y_{\amaj} + h \cdot y_{\amin}, 
\end{array}\label{coords.eqn}
\end{equation}
where $w$ and $h$ are width and height respectively. The width and height will typically take into account the resolution of the major axes, that is, the minimum difference between consecutive locations.

Computing the coordinates for star glyphs is achieved by using polar coordinates,  $r=y_{\amin}$, and $\theta=x_{\amin}$, both scaled to be between 0 and 1. Then the glyph coordinates are given by 
\begin{equation}
\begin{array}{lll}
x_\text{ coordinate}&=& x_{\amaj} + \frac{w \cdot r}{2} \sin(2 \pi \theta) \\
y_\text{ coordinate}&=& y_{\amaj} + \frac{h \cdot r}{2} \cos(2 \pi \theta).
\end{array}\label{coords.polar.eqn}
\end{equation}
Note that, this is is a non-standard conversion to polar coordinates, but this conversion provides a time line that starts at 12 o'clock and wraps around the origin in a clockwise fashion.

@@

% Show the glyph-map for de-seasonalized data, and describe what we
% see differently from the facetted map

%\begin{figure}[htbp]
%  \centering
%  \includegraphics[width=1\linewidth]{gistemp-raw}%

%  \includegraphics[width=1\linewidth]{gistemp-polar-raw}

%  \caption{Glyph-maps of gridded raw temperature anomaly data}
%  \label{fig:ref-gistemp}
%\end{figure}

@@

@@

The glyph-map can be considered to be small multiples as named by \citet{tufte:2001}, in this case, small multiples of time series. By laying the time series out on the spatial grid, the main task expected of the viewer is comparing the temporal patterns from one location to the next. In contrast, in the small multiples used in the facetted colored map, the viewer is expected to compare the patterns in the maps from one year to the next, for corresponding months. (See \citet{carr:1999} for a discussion on perceptual grouping.) That is, compare and contrast maps with facetted colored map, or compare and contrast time series using the glyph-map. One draw-back of the glyph-maps, particularly if linear trend is displayed as an icon, is they may suffer from the Z\"ollner Illusion \citep{Zollner} which is visible in Figure \ref{fig:gistemp-pred}. On the other hand, comparing images require that we play a spot the difference game, which is perceptually difficult \citet{busey}.

@@

\begin{figure*}[htp]
\centerline{\includegraphics[width=4in]{nasa-deseas-glyph.png}}
\caption{Glyph-map of de-seasonalized temperature, to compare with Figure \ref{fig:facetted-map}. }
\label{fig:nasa-deseas-glyph}
\end{figure*}

@@

Structure in temporal data is composed of different parts: global trend, seasonality and error. Extracting each of these for visualization improves the focus on different aspects of the time series. A generalized additive model \citep{wood:2006} with a seasonal term and a long-term smooth term can be useful. This is revealing for smooth models of surface temperature, shown in Figure~\ref{fig:scaling}. Each location is modeled using a generalized additive model \citep{wood:2006} with a seasonal term and a long-term smooth term. The three plots in the figure show the long-term smooth term. %Because some locations are uniformly warmer than others, the impact of changes across time are diminished. In the center plot, each location is scaled by dividing out the maximum, and on the right, each location is scaled to range $[0, 1]$. These emphasize individual changes but lose the ability to make global comparisons, and need caution when interpreting - a small absolute change gets blown up into a large relative change.
 %The three plots in the figure show the long-term smooth term. Because some locations are uniformly warmer than others, the impact of changes across time are diminished.

@@

\section{Long time series}

>>>>

Climate data is composed of measurements on variables such as temperature, precipitation, and winds that have been collected from a spatial and temporal context. The classic display of this type of is to use color or grey scale overlaid on geographic location, typically including a map. When measurements are made at different time periods it is common to display the data as small multiples, such as in Figure~\ref{fig:facetted-map}. In this plot, a separate map is drawn for each month (columns) and year (rows), over a 6 year period of remotely sensed temperature data above Central America \citet{murrell:2010}. Color is used to display the de-seasonalized temperature measurements, with red being high values and blue being low values. The most noticeable feature in this display is the strong red in the equatorial region beginning in mid-1997 and tapering out during 1998. This is the pattern of an El Ni\~no event, a major temperature anomaly. With some work other more localized patterns can be perceived, for example, cooler over the land mass in the early year, but warmer in the latter years.

Reading this plot is cognitively challenging: the reader must play ``spot the difference'' from one small image to another. Large structures such as the El Ni\~no event are clear but it is very difficult to glue the images together to read off the trend over the years or to notice local deviations. Rendering these small multiples as a movie can help with picking up strong trends, but small trends still fail to draw readers' attention and escape unnoticed \citep{simons:gradual}.

\begin{figure*}[htbp]
  \centering
  \includegraphics[width=5.5in]{nasa-colored-map.png}
  \caption{Facetted map of de-seasonalized temperature. The dominant feature is the El Ni\~no warming in the southern equatorial region in the last half of 1997 and first half of 1998. Smaller features are only noticeable on closer inspection, or if pointed out: such as the relative warming on the mountain regions in south and north America in the later years.}
  \label{fig:facetted-map}
\end{figure*}

@@

In this paper we develop a different type of display that resolves these problems: the \emph{glyph-map}. The glyph-map uses a small glyph, or icon, to represent multiple values at each location. Glyph maps are an adaptation of glyphs, tools developed for display multivariate data, including the star plots of \citet{mayr:1877}, the semi-graphic displays of \citet{anderson:1960}, and the infamous Chernoff faces \citep{chernoff:1973}. An glyph is produced by mapping each variable to some graphical feature. Figure~\ref{fig:nasa-deseas-glyph} shows the glyph-map equivalent of Figure~\ref{fig:facetted-map}: each location is represented by a tiny time series glyph. The glyph-map can also be thought of as a small multiple display \citet{tufte:2001} of time series.

\begin{figure*}[htbp]
  \centering
  \includegraphics[width=3.5in]{nasa-deseas-glyph.png}
  \caption{Glyph-map of de-seasonalized temperature, to compare with Figure \ref{fig:facetted-map}. }
  \label{fig:nasa-deseas-glyph}
\end{figure*}

One problem with glyph displays of multivariate data is that there is no natural ordering of each glyph, or of the mapping between variables and glyph properties. This leads to a combinatorial explosion of possibilities which may have dramatically different perceptual properties. Despite some proposed remedies \citep{kleiner:1981,hurley:2010}, this problem has prevented the widespread adoption of glyph displays. Using glyphs for space-time data resolves this problem: we place glyphs according to the measurement location of the data, and we order variables in time. Additionally, climate data usually has high-correlations between nearby locations and times. This imposes a certain degree of smoothness that gives the plot an appearance of texturing on a landscape, making it easier to digest patterns.
Others \citet{pickett:1988} have used glyphs on maps to display multivariate spatial data. This area of work has morphed into the field of metaphorical data displays, which creates abstract landscapes of spatial data, a digression from glyph-maps. \citet{gribov:2006} describes the use of glyph-maps for multivariate data also, with emphasis on the graphics software Gauguin.

We focus our attention on two glyphs: lines and stars. Figure~\ref{fig:templates} displays 12 iconic time series shapes with line and star glyphs. The data underlying each glyph has 36 time variables. The star glyphs is formed by considering the 36 axes radiating from a common midpoint, and the data values for the row are plotted on each axis relative to the locations of the minimum and maximum of the variable.

@@

In this paper we describe the algorithm for creating glyphs of temporal data on maps, and the various modifications that can enable different features in the data to be detected. Section~\ref{sec:glyph-map} describes the use of glyphs on maps, and Section~\ref{sec:construction} explains how to construct glyph-maps. Section~\ref{sec:reference} discusses the use of reference lines and boxes, and Section~\ref{sec:scale} visits issues of scaling values for different information purposes. Generalizations to different types of icons are described in Section~\ref{sec:scatter}. Glyph-maps work nicely with gridded data, but many spatiotemporal data sets have irregular spatial locations. Possible adjustments are discussed in Section \ref{sec:irregular}.

To demonstrate the glyph-map we will use three datasets:

\begin{itemize}

  \item The ASA 2009 data expo data which consists of monthly observations of
  several atmospheric variables from the International Satellite Cloud
  Climatology Project. The dataset includes observations over 72 months
  (1995--2000) on a 24 x 24 grid (576 locations) stretching from
  $113.75^{\circ}$W to $56.25^{\circ}$W longitude and $21.25^{\circ}$S to
  $36.25{^\circ}$N latitude.

  \item GISTEMP surface temperature data provided by the Earth System Research
  Laboratory, Physical Sciences Division, National Oceanic and Atmospheric
  Administration. This data is provided on $2^{\circ}$ x $2^{\circ}$ grid over
  the continental United States (XXX locations), measured monthly (???) from
  XXX to YYY (ZZZ time points in total). Ground station data was
  de-seasonalized, differenced from ***, and spatially averaged to obtain
  gridded measurements

  \item USHCN (Version 2) ground station network.

\end{itemize}

All datasets and R code used to produce the plots are available in the online supplementary materials.

\section{Glyph plots for large data}
\label{sec:construction}

Figure~\ref{fig:gistemp-pred} shows the a glyph plot for a dataset that represents a more typical size of climate data, the GISTEMP data. Star glyphs at each location represent temperature from 1880--2011 across the USA. The major feature to notice is the pattern of missing values, glyphs that do not complete the full circle. The high noon position of the circle is the first time point, Jan 1880, and it is this early time period that has missing values are a lot of locations. Another important feature is variability, indicated by the thickness of the circle. Some locations, particularly in the north western mountain region, have big variability in temperature, both low minima, and high maxima. At these locations there are large fluctuations from the long term mean.

@@

Large datasets often required the display of summaries other than the raw data. Structure in temporal data is composed of different parts: global trend, seasonality and error. Extracting each of these for visualization improves the focus on different aspects of the time series. Figure~\ref{fig:gistemp-pred} shows the time series icons for each spatial location of predicted temperature for 1950--2010 across the USA (limited time period used due to missing data). Most locations observe an increasing temperature, to varying degrees, over this time period. A few locations show a decline. These locations tend to be different from their neighbors, which might suggest isolated data problems. There is a difference in the temperature increase at different regions: steeper inclines can be seen in the north and western mountain region and across the Great Lakes, but inclines are flatter in the southeastern USA.

\begin{figure}[htbp]
  \centering
  \includegraphics[width=1\linewidth]{gistemp-pred}%
  %\includegraphics[width=1\linewidth]{gistemp-polar-pred}

%The most seasonality occurs in north American. Some seasonal effects can also be seen in the Caribbean and in south American locations. Around the equator the temperatures are relatively warm for all seasons. Instead of radially displaying the temperature a glyph can be a regular time series, time displayed horizontally and temperature vertically (right plot). The structure perceived is different with this change in glyph type: locations at the equator have flat series, on land more varied temperatures.

@@

Glyph-maps can be generated with existing graphics software after performing a simple pre-processing step, making them easily accessible to a wide audience. Creating a glyph-map, involves recognizing that they combine two structural components of the data: spatial location and data values. The spatial location is the major positioning component, and the data values are minor adjustments to these positions. For spatiotemporal data, the major axes are latitude ($y_{\amaj}$) and longitude ($x_{\amaj}$), and the minor axes are time ($x_{\amin}$) and some measurement ($y_{\amin}$), for example, temperature, or predicted temperature. The minor axes are rescaled to have range $[0, 1]$, and further adjusted to fit into the resolution of the spatial coordinates. The final values to be plotted are a simple linear combination, as given by

\begin{equation}
  \begin{array}{lll}
  x_\text{ coordinate}&=& x_{\amaj} + w \cdot x_{\amin}\\
  y_\text{ coordinate}&=& y_{\amaj} + h \cdot y_{\amin}, 
  \end{array}
  \label{coords.eqn}
\end{equation}

\noindent where $w$ and $h$ are width and height respectively. The width and height will typically take into account the resolution of the major axes, that is, the minimum difference between consecutive locations.

Computing the coordinates for star glyphs is achieved with a simple polar transformation:

\begin{equation}
  \begin{array}{lll}
  x_\text{ coordinate}&=& x_{\amaj} + y_{\amin} \cdot \frac{w}{2} \sin(2 \pi x_{\amin}) \\
  y_\text{ coordinate}&=& y_{\amaj} + y_{\amin} \cdot \frac{h}{2} \cos(2 \pi x_{\amin}).
  \end{array}
  \label{coords.polar.eqn}
\end{equation}

\noindent Note that this is is a non-standard conversion to polar coordinates, but it provides a timeline that starts at 12 o'clock and travels clockwise.

@@

@@

% From a cognitive perspective, this is a very hard task to do. It falls into the framework of {\it change blindness} (see e.g. \citet{healey:2011} and often leads to failure without readers realizing, in particular for small areas of change or subtle changes.

@@

By laying the time series out on the spatial grid, the main task expected of the viewer is comparing the temporal patterns from one location to the next. In contrast, in the small multiples used in the facetted colored map, the viewer is expected to compare the patterns in the maps from one year to the next, for corresponding months. (See \citet{carr:1999} for a discussion on perceptual grouping.) That is, compare and contrast maps with facetted colored map, or compare and contrast time series using the glyph-map. One draw-back of the glyph-maps, particularly if linear trend is displayed as an icon, is they may suffer from the Z\"ollner Illusion \citep{Zollner} which is visible in Figure \ref{fig:gistemp-pred}. On the other hand, comparing images require that we play a spot the difference game, which is perceptually difficult \citet{busey}.

@@

@@

A generalized additive model \citep{wood:2006} with a seasonal term and a long-term smooth term can be useful. This is revealing for smooth models of surface temperature, shown in Figure~\ref{fig:scaling}. Each location is modeled using a generalized additive model \citep{wood:2006} with a seasonal term and a long-term smooth term. The three plots in the figure show the long-term smooth term. %Because some locations are uniformly warmer than others, the impact of changes across time are diminished. In the center plot, each location is scaled by dividing out the maximum, and on the right, each location is scaled to range $[0, 1]$. These emphasize individual changes but lose the ability to make global comparisons, and need caution when interpreting - a small absolute change gets blown up into a large relative change.
 %The three plots in the figure show the long-term smooth term. Because some locations are uniformly warmer than others, the impact of changes across time are diminished.

@@
